---
title: 'Intro to R'
author: "Juan Vélez"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  ioslides_presentation:
    css: default2.css
    widescreen: true
---

# Why learn R?


## Advantages of using R

1. Free
2. Infinitely more intuitive and regular than Stata
3. Can be used as a calculator
4. You can have many objects in your environment
6. Integrated version control
5. Potential to understand better what you are doing
5. Makes reproducible research easy
5. Huge community willing to help
7. Kickass presentations like this one



## Disadvantages

1. First time getting clustered standard errors (2 or more way) might be tough
3. Marginally slower than Stata for regressions with many fixed-effects


# Interacting with R(-studio)

## Basics

- R is the program actually doing all the work
- R-Studio is an interface to improve the experience  
- R-Studio layout:  
    - Interactive console (left)
    - Environment/History (tabbed in upper right)
    - Files/Plots/Packages/Help/Viewer (tabbed in lower right)

- You can easily get Help by typing 1 or 2 interrogation marks in the console


## R as calculator

The simplest thing you could do in R is arithmetic and logic

```{r, results='asis'}
100 + 4
```

If you type an incomplete statement, R will wait for you. The '+' sign indicates R is waiting for you.


## R as calculator (II)

When using R as calculator the order of operations is (highest to lowest precedence)

- Parentheses: `(` `)`
- Exponents: `^` or `**`
- Divide: `/`
- Multiply: `*`
- Add: `+`
- Subtract: `-`

```{r, results='asis'}
3 + 5 * 2
(3 + 5) * 2
```


## R as calculator (III)

- R expresses small numbers in scientific notation

```{r, results='asis'}
2/10000

```

- You can do it too

```{r, results='asis'}
5e3 # note the lack of minus sign here
```


- Math functions

```{r, results='asis'}
28 %% 6
```


## Comparing things

R can also be used to compare things


```{r, results='asis'}
1 == 1 # note the double equal (is equal to)
```

```{r, results='asis'}
1 != 2 # note the exclamation mark (is not equal to)
```

```{r, results='asis'}
1 < 2 # less than
```

```{r, results='asis'}
1 <= 2 # less than or equal
```

## Comparing things (II)

```{r,results='asis'}
4 > 2 #greater than

```

```{r,results='asis'}
4 >= 4 #greater than or equal

```

<br>
Some funny things

```{r,results='asis'}
'hola' > 4e100
T > -5
```


## Variable assignment

To store values, R prefers the operator `<-`

```{r, results='asis'}
y <- 1/4
y
```

One nice feature of R: evaluations happens before assignment

```{r, results='asis'}
A <- 100
A <- A+1
A
```

## Vectorization

R building block is the vector, for instance

```{r, results='asis'}
1:5
2^(1:5)
x <- 1:5
2^x
```

## Managing your environment

If you want a list of all the objects in your environment

```{r, results as is}
ls()
```

If you want to remove a variable

```{r, results='asis'}
rm(A)
ls()
```
If you want to remove them all

```{r}
rm(list=ls())
ls()
```

## R packages

It is possible to add functions to R by writing or obtaining packages. You can

- See installed packages by typing `installed.packages()`
- Install packages by typing `install.packages("packagename")`, where `packagename` is the package name, in quotes.
- Update installed packages by typing `update.packages()`
- Remove a package with `remove.packages("packagename")`
- Make a package available for use with `library(packagename)`


## Challenges

- What will be the value of each variable after each statement in the following program?
```{r, results='asis'}
mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age - 20
``` 
- Run the code from the previous challenge, and write a command to compare mass to age. Is mass larger than age?
- Clean up your working environment by deleting the mass and age variables.
- Install the following packages: `ggplot2`, `dplyr`, `gapminder`


# Creating a self contained project

## A folder for the project

1. Click the “File” menu button, then “New Project”.
2. Click “New Directory”.
3. Click “Empty Project”.
4. Type in the name of the directory to store your project, e.g. “my_project”.
6. Click the “Create Project” button.

## Data to work later

Download the Gapminder data from the link provided in the etherpad.

- Download the file (CTRL + S, right mouse click -> “Save as”, or File -> “Save page as”)
- Make sure it’s saved under the name `gapminder-FiveYearData.csv`
- Save the file in your project folder.

We will load and inspect these data later.



## Finding help

R, and every package, provide help files for functions

```{r, eval=F}
?function_name
help(function_name)
```
To find help about special operators, quotes are required

```{r, eval=F}
?'%%'
```

Challenge: Look at the help for the c function. What kind of vector do you expect you will create if you evaluate the following:

```{r, eval=F}
c(1, 2, 3)
c('d', 'e', 'f')
c(1, 2, 'f')`
```


## Data types

R deals with numerous data types. The basic four are:

- Decimals values like `4.5` are called numerics.  

- Natural numbers like `4` are called integers. Integers are also numerics. 

- Boolean values (`TRUE` or `FALSE`) are called logical.  

- Text (or string) values are called characters.  


## Data structures

R can handle many types of structures. We will focus on 

- Vectors: have no dimension, all elements of the same type
- Matrices: have dimension, all elements of the same type
- Data frames: have dimensions, elements can be of different type as long as they are the same length
- Lists: super flexible, you can put anything in a list



## Going to Las Vegas (Challenge) {.smaller}

- Assign the value `"Go!"` to the variable `vegas`

- Create 2 vectors containing your winnings and loses from poker and roulette

    -For `poker.vector`:
    
        - On Monday you won $140  
        - Tuesday you lost $50  
        - Wednesday you won $20  
        - Thursday you lost $120  
        - Friday you won $240 

    -For `roulette.vector`:
    
        - On Monday you lost $24  
        - Tuesday you lost $50  
        - Wednesday you won $100  
        - Thursday you lost $350  
        - Friday you won $10  

## Going to Las Vegas (Challenge) (II) {.smaller}

Elements in vectors can be named

```{r, eval=F}
some.vector <- c("Pepito Perez", "Futbolista")
names(some.vector) <- c("Nombre", "Profesión")
```

Following the example, name the elements of the roulette winnings
```{r, eval=F}
names(poker.vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roullete.vector) <- 
```

- Create a vector with the net winnings of each day. Name it `total.daily`
- Create a named vector telling which days you made winnings. Name it `days.won`
- Calculate the total winnings from the 5 days in Las Vegas. Name it `total.week`.


## Going to Las Vegas (III) {.smaller}

Slicing in R is really simple: ff you want the first three elements of a vector just use '[]' and the appropriate indexes (from 1 to 3)

```{r, eval = F}
total.daily[1:3] # gets the daily wins from Mon-Wed

```

You can also perform operations and apply functions to parts of vectors

```{r, eval=F}
mean(total.daily) # gives the mean for the whole week
mean(total.daily[4,5]) # gives the mean from Thu-Fri
```


## Matrices {.smaller}

Think of matrices as vector with dimensions or a stack of vectors (of the same dimension)

```{r}

matrix(1:9, nrow = 3, ncol = 3)

```

By default the function 'matrix' fills out the blank matrix by column

```{r}

matrix(1:9, byrow = T, ncol = 3)
```


## Matrices challenge {.smaller}

- The following vectors are the number of copies sold for 3 of Miles Davis' albums. 

- The first number of each vector is the US sales and the second is the rest of the world.

```{r, eval = F}
kind.of.blue <- c(460.998, 314.4)
sketches <- c(290.475, 247.900)
miles.ahead <- c(309.306, 165.8)
```

- Combine these three vectors into one single vector called 'all.sales'.

- Using the vector 'all.sales', create a matrix called 'miles.sales' that has US and rest of the world sales as columns.

- Using help or Google, label the columns and rows of the matrix properly.


## Selection of matrix elements {.smaller}

We can naturally slice a matrix by using the `[]` operator

```{r, eval=F}
some.matrix[3,6] # gets the element in the 3rd row and 6th column
some.matrix[,6] # gets all the elements in the 6th column
some.matrix[3,] # gets all the elements in the 3rd row
```


Challenge

- Create a 50 by 20  matrix with the numbers from 1 to 1000 (the first column is the numbers from 1-50). Name it `my.matrix`
- Extract the following matrix from `my.matrix`

```{r, echo=F}
my.matrix <- matrix(1:1000, nrow = 50, ncol = 20)
 my.matrix[39:45, 1:4]
```



## Arithmetic with matrices (I) {.smaller}

Just as vectors, the normal operators perform element-wise operations

```{r}
a <- matrix(5:8, nrow = 2, ncol = 2)
b <- matrix(1:4, byrow = T, ncol = 2)
c <- a * b
```

Scalars work the same way

```{r}
d <- matrix(1:9, nrow = 3)
pi * d
```


## Arithmetic with matrices (II) {.smaller}

If we want  to use standard matrix multiplication we need to use the operator `%*%`

```{r}
e <- matrix(1:6, nrow = 3, ncol = 2)
f <- matrix(11:20, nrow = 2, ncol = 5)
g <- e %*% f
```


For inverse we use the function `solve()`

```{r}
h <- matrix(rnorm(9), nrow = 3)
h.inv <- solve(h)
```

for transpose the function `t()`

```{r}
i <- matrix(21:30, byrow = T, ncol = 2)
j <- t(i)
```


## Functions and matrices

Applying functions to columns or rows is as easy as


```{r}
k <- matrix(rnorm(100), byrow = T, nrow = 10)

column.mean <- mean(k[,1])
row.mean <- mean(k[4,])

```


Challenge:

- Calculate the mean and standard deviation for Miles Davis' sales for US and non-US markets separately.


## Factors {.smaller}

Are the categorical variables in R. It is not just a name. They are treated particularly. They are extremely important for research in Economics.

```{r}
genre <- c('bebop', 'modal', 'cool', 'modal', 'cool', 'cool')
genre <- factor(genre)
```

Some factors are naturally ordered (e.g days of the week) while some others aren't

```{r}
fruits <- c('Lulo', 'Guama', 'Carambolo', 'Papaya', 'Guama')
fruits <- factor(fruits)

temperatures <- c('High', 'Medium', 'High', 'High', 'Low', 'Low')
temperatures <- factor(temperatures, ordered = T, levels = c('Low', 'Medium', 'High'))

```

And you can always change the name of your levels

```{r}
levels(temperatures) <- c('L', 'M', 'H')
```



## Data frames

- This is the most typical data structure that economists work with. 
- Data in data frames can be of numerous types (but same length)
- A typical example is household surveys    
    - name (`character`)
    - age (`numeric`)
    - married (`logical`)

- Some ways of getting to know your DF
```{r, eval=F}
mtcars # pre-loaded data frame always available in R
head(mtcars)
tail(mtcars)
str(mtcars)
```


## Creating dataframes 

- Most of the time you will read a csv table to create DF.
- Sometimes, though, you have to create them

```{r}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter",
             "Saturn", "Uranus", "Neptune")
type <- c("Terrestrial", "Terrestrial", "Terrestrial", "Terrestrial", 
          "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

```

- Challenge:
    - Create a data frame with the previous vectors. Call it `planets.df`. 
    - Use the function `str()` to analyze the structure of `planets.df`.

## Slicing data frames

- Similar to matrices, the `[]` can be used

```{r, eval=F}
df[1,2] # selects the first elemnt of column 2
df[1:3, 2:8] # selects the first 3 elemnts of columns 2 to 8
df[1,] # what would this select?
```

- Challenge:  
    - Print out the value of the diameter of Mercury
    - Print out all data on Mars

> - names of columns can also be used to slice
    - Challenge:  Select the diameter of the furthest planets
    

## Slicing data frames: columns

There are three ways of selecting columns

```{r, eval=F}
planets.df[,3]
planets.df[,'diameter']
planets.df$diameter
```

The last way is clearly less verbose. more concise.


# Data frame: conditional selecting

- Suppose we want to work only with planets that have rings. 

```{r, eval=F}
has.rings <- planets.df$rings
planets.with.rings.df <- planets.df[rings,]
```

- the function `subset()` can be used as a shorthand

```{r, eval=F}
planets.with.rings.df2 <- subset(planets.df, subset = rings)

```

> - Challenge:  
      - Create a dataframe with the planets whose diameter is less than 1.




## Lists

- Think of lists as rucksacks or backpacks. At the end you want to put many things in one container.


```{r, eval=F}
my.list <- list(a,b,c,d, genre.vector, planets.df)
my.list
```

- If you want the elements of the list to have names


```{r, eval=F}
my.named.list <- list(some.matrix = b, some.df = planets.df, my.age = 36)
```



## Working with Dplyr and Magrittr: prelude

- Let's open a new R-script. Name it `my.script.R`.
- Using the function `read.csv()` read the csv file with the gapminder data.
- This data is clean. It usually isn't.
- load the libraries `dplyr` and `magrittr`
> - Challenge: Describe the data frame (dimensions and data types)



## dplyr

- It was created with data frames in mind
- It is awesome to manipulate data in many, many useful ways
- the names of the functions are the same verbs we'd use to describe what we want  
    - `filter()` (and `slice()`)
    - `arrange()`
    - `Select()` (and `rename()`)
    - `distinct()`
    - `mutate()` (and `transmute()`)
    - `summarize`
    - `sample_n()` (and `sample_frac()`)


## magrittr

- It is meant to write code in a 'pipeline' way
- Code written using `magrittr` is more readable 
- It is probably clearer with examples


## Data frame challenges

Do the following using the gapminder data

- Remove data from Africa and from North Korea (`filter()`)
- keep only data from 2007
- Create a column with the gdp of each country (`mutate()`)
- Group the data by continent (`group_by()`)
- create a variable with the total population of the continent(call it `continent.pop`).
- Create a variable with the number of countries per continent (`n()`) call it `num.country`
- Summarize the data frame at continent level (the result should have 4 continents and 4 variables)


```{r, echo=F, eval=F}
df.gr <- df %>%
  filter(continent != 'Africa', country != 'Korea Dem. Rep.') %>%
  filter(year == '2007') %>%
  mutate(gdp.country = gdpPercap*pop) %>%
  group_by(continent) %>%
  mutate(continent.pop = sum(pop),
         num.country = n()) %>%
  summarize(avg.pop.cont = mean(continent.pop),
            num.country = max(num.country),
            gdp.cont = sum(gdp.country))
```



