---
title: 'Intro to R'
author: "Juan Vélez"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  ioslides_presentation:
    css: default2.css
    widescreen: true
---

# Why learn R?


## Advantages of using R

1. Free
2. Infinitely more intuitive and regular than Stata
3. Can be used as a calculator
4. You can have many objects in your environment
6. Integrated version control
5. Potential to understand better what you are doing
5. Makes reproducible reasearch easy
5. Huge community willing to help
7. Kickass presentations like this one



## Disadvantages

1. First time getting clustered standard errors (2 or more way) might be tough
3. Marginally slower than Stata for regressions with many fixed-effects


# Interacting with R(-studio)

## Basics

- R is the program actually doing all the work
- R-Studio is an interface to improve the experience  
- R-Studio layout:  
    - Interactive console (left)
    - Environment/History (tabbed in upper right)
    - Files/Plots/Packages/Help/Viewer (tabbed in lower right)

- You can easily get Help by typing 1 or 2 interrogation marks in the console


## R as calculator

The simples thing you could do in R is arithmetic and logic

```{r, results='asis'}
100 + 4
```

If you type an incomplete statement, R will wait for you. The '+' sign indicates R is waiting for you.


## R as calculator (II)

When using R as calculator the order of operations is (highest to lowest precedence)

- Parentheses: `(` `)`
- Exponents: `^` or `**`
- Divide: `/`
- Multiply: `*`
- Add: `+`
- Subtract: `-`

```{r, results='asis'}
3 + 5 * 2
(3 + 5) * 2
```


## R as calculator (III)

- R expresses small numbers in scientific notation

```{r, results='asis'}
2/10000

```

- You can do it too

```{r, results='asis'}
5e3 # note the lack of minus sign here
```


- Math functions

```{r, results='asis'}
28 %% 6
```


## Comparing things

R can also be used to compare things


```{r, results='asis'}
1 == 1 # note the double equal (is equal to)
```

```{r, results='asis'}
1 != 2 # note the exclamation mark (is not equal to)
```

```{r, results='asis'}
1 < 2 # less than
```

```{r, results='asis'}
1 <= 2 # less than or equal
```

## Comparing things (II)

```{r,results='asis'}
4 > 2 #greater than

```

```{r,results='asis'}
4 >= 4 #greater than or equal

```

<br>
Some funny things

```{r,results='asis'}
'hola' > 4e100
T > -5
```


## Variable assignment

To store values, R prefers the operator `<-`

```{r, results='asis'}
y <- 1/4
y
```

One nice feature of R: evaluations happens before assignment

```{r, results='asis'}
A <- 100
A <- A+1
A
```

## Vectorization

R building block is the vector, for instance

```{r, results='asis'}
1:5
2^(1:5)
x <- 1:5
2^x
```

## Managing your environment

If you want a list of all the objects in your environment

```{r, results as is}
ls()
```

If you want to remove a variable

```{r, results='asis'}
rm(A)
ls()
```
If you want to remove them all

```{r}
rm(list=ls())
ls()
```

## R packages

It is possible to add functions to R by writing or obtaining packages. You can

- See installed packages by typing `installed.packages()`
- Install packages by typing `install.packages("packagename")`, where `packagename` is the package name, in quotes.
- Update installed packages by typing `update.packages()`
- Remove a package with `remove.packages("packagename")`
- Make a package available for use with `library(packagename)`


## Challenges

- What will be the value of each variable after each statement in the following program?
```{r, results='asis'}
mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age - 20
``` 
- Run the code from the previous challenge, and write a command to compare mass to age. Is mass larger than age?
- Clean up your working environment by deleting the mass and age variables.
- Install the following packages: `ggplot2`, `dplyr`, `gapminder`


# Creating a self contained project

## A folder for the project

1. Click the “File” menu button, then “New Project”.
2. Click “New Directory”.
3. Click “Empty Project”.
4. Type in the name of the directory to store your project, e.g. “my_project”.
6. Click the “Create Project” button.

## Data to work later

Download the Gapminder data from the link provided in the etherpad.

- Download the file (CTRL + S, right mouse click -> “Save as”, or File -> “Save page as”)
- Make sure it’s saved under the name `gapminder-FiveYearData.csv`
- Save the file in your project folder.

We will load and inspect these data later.



## Finding help

R, and every package, provide help files for functions

```{r, eval=F}
?function_name
help(function_name)
```
To find help about special operators, quotes are required

```{r, eval=F}
?'%%'
```

Challenge: Look at the help for the c function. What kind of vector do you expect you will create if you evaluate the following:

```{r, eval=F}
c(1, 2, 3)
c('d', 'e', 'f')
c(1, 2, 'f')`
```


## Data types

R deals with numerous data types. The basic four are:

- Decimals values like `4.5` are called numerics.  

- Natural numbers like `4` are called integers. Integers are also numerics. 

- Boolean values (`TRUE` or `FALSE`) are called logical.  

- Text (or string) values are called characters.  


## Data structures

R can handle many types of structures. We will focus on 

- Vectors: have no dimension, all elements of the same type
- Matrices: have dimension, all elements of the same type
- Data frames: have dimensions, elements can be of different type as long as they are the same length
- Lists: super flexible, you can put anything in a list



## Going to Las Vegas (Challenge) {.smaller}

- Assign the value `"Go!"` to the variable `vegas`

- Create 2 vectors containing your winnings and loses from poker and roulette

    -For `poker.vector`:
    
        - On Monday you won $140  
        - Tuesday you lost $50  
        - Wednesday you won $20  
        - Thursday you lost $120  
        - Friday you won $240 

    -For `roulette.vector`:
    
        - On Monday you lost $24  
        - Tuesday you lost $50  
        - Wednesday you won $100  
        - Thursday you lost $350  
        - Friday you won $10  

## Going to Las Vegas (Challenge) (II) {.smaller}

Elements in vectors can be named

```{r, eval=F}
some.vector <- c("Pepito Perez", "Futbolista")
names(some.vector) <- c("Nombre", "Profesión")
```

Following the example, name the elements of the roulette winnings
```{r, eval=F}
names(poker.vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roullete.vector) <- 
```

- Create a vector with the net winnings of each day. Name it `total.daily`
- Create a named vector telling which days you made winnings. Name it `days.won`
- Calculate the total winnings from the 5 days in Las Vegas. Name it `total.week`.


## Going to Las Vegas (III) {.smaller}

Slicing in R is really simple: ff you want the first three elements of a vector just use '[]' and the apropriate indexes (from 1 to 3)

```{r, eval = F}
total.daily[1:3] # gets the daily wins from Mon-Wed

```

You can also perform operations and apply functions to parts of vectors

```{r, eval=F}
mean(total.daily) # gives the mean for the whole week
mean(total.daily[4,5]) # gives the mean from Thu-Fri
```


## Matrices {.smaller}

Think of matrices as vector with dimensions or a stack of vectors (of the same dimension)

```{r}

matrix(1:9, nrow = 3, ncol = 3)

```

By default the function 'matrix' fills out the blank matrix by column

```{r}

matrix(1:9, byrow = T, ncol = 3)
```


## Matrices challenge {.smaller}

- The following vectors are the number of copies sold for 3 of Miles Davis' albums. 

- The first number of each vector is the US sales and the second is the rest of the world.

```{r, eval = F}
kind.of.blue <- c(460.998, 314.4)
sketches <- c(290.475, 247.900)
miles.ahead <- c(309.306, 165.8)
```

- Combine these three vectors into one single vector called 'all.sales'.

- Using the vector 'all.sales', create a matrix called 'miles.sales' that has US and rest of the world sales as columns.

- Using help or google, label the columns and rows of the matrix properly.
